---
title: "ANPC SkylineR and Lipid ExploreR Quality Control Report"
output: html_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```

Thank you for using LGW SkylineR and Lipid ExploreR. Your lipidomics quality control evaluation report will now be produced.  

# 1. Stable isotope lable (SIL) internal standard (IS) quality control check

The first QC check summed the peak areas from all stable isotope labeled (SIL) internal standards (IS) to identify those samples that either:  

1 - contained low/no internal standard. Suggesting a preparation error where low volume of the SIL IS had been added to the well  
2 - contained excessive internal standard. Suggesting a preparation error where a droplet may have entered the well increasing the volume of the SIL IS  



```{r SIL QC cehck, echo=FALSE, message=FALSE, fig.width=10, fig.height=4}
sil_check_p 

print(paste(nrow(sil_qc_fail_samples), " study samples and ", nrow(sil_qc_fail_ltr), " LTR samples were removed from the dataset during this check", sep = ""))

knitr::kable(sil_qc_fail)
```


# 2. Total ion count (TIC) quality control check

The second QC check summed the total ion count of all lipid targets to identify those samples that:  

1 - contained low/no signal. Suggesting a preparation error where low volume of the sample had been added to the well  


```{r TIC QC check, echo=FALSE, message=FALSE, fig.width=10, fig.height=4}
tic_check_p 

print(paste(nrow(tic_qc_fail_samples), " study samples and ", nrow(tic_qc_fail_ltr), " LTR samples were removed from the dataset during this check", sep = ""))

knitr::kable(tic_qc_fail)
```

# 3. Target lipid feature quality control

The QC checks now switched to the lipid mass spectrometry targets. Lipids were filtered out due to a user defined singal intensity.

```{r lipid target intensity filter, echo=FALSE, message=FALSE, fig.width=10, fig.height=4}


if(intensity_threshold_ltr == "LTR"){
  print(paste("A lipid feature was kept in the dataset if it was present in ", intensity_threshold_percentage, " % of ",  intensity_threshold_ltr, " samples with a signal intensity (peak height) greater than ", intensity_threshold, "counts", sep = ""))
}

if(intensity_threshold_ltr == "samples"){
  print(paste("A lipid feature was kept in the dataset if it was present in ", intensity_threshold_percentage, " % of study",  intensity_threshold_ltr, " (not LTR) with a signal intensity (peak height) greater than ", intensity_threshold, "counts", sep = ""))
}

if(intensity_threshold_ltr == "both"){
  print(paste("A lipid feature was kept in the dataset if it was present in ", intensity_threshold_percentage, " % of both study samples and LTR samples with a signal intensity (peak height) greater than ", intensity_threshold, "counts", sep = ""))
}

print(paste(length(lipid_intensity_filter_fail),"lipids were removed from the dataset as they did not meet these criteria"))


```

# 4. Internal standard feature normalization

Following normalization with internal standard (pre-defined by the user) features were removed if the percentage relative standard deviation (%RSD) was greater than 30%

```{r IS LTR filter, echo=FALSE, message=FALSE, fig.width=10, fig.height=4}

print(paste("Total number of feature ratios with with an LTR RSD of <30% =", length(which(ltr_rsd$RSD < 30))))
print(paste("Total number of feature ratios with with an LTR RSD of <20% =", length(which(ltr_rsd$RSD < 20))))
print(paste("Total number of feature ratios with with an LTR RSD of <15% =", length(which(ltr_rsd$RSD < 15))))
print(paste("Total number of feature ratios with with an LTR RSD of <10% =", length(which(ltr_rsd$RSD < 10))))


```

#5. Visualisation of normalised data
normalized_check_class_p

```{r Visualisation of normalised data, echo=FALSE, message=FALSE, fig.width=10, fig.height=8}

print(paste("Plot of lipid feature ratios normalised by internal standards. This plot can be used to look for trends in the data such as run order and plate/batch effects"))
normalized_check_p

print(paste("Plot of lipid feature ratios normalised by internal standards - seperated into lipid classes. This plot can be used to look for trends in the data such as run order and plate/batch effeects"))
normalized_check_class_p
```

# 6. PCA plot to visualize final dataset (LTR and samples)


```{r PCA for the report, echo=FALSE, message=FALSE, fig.width=10, fig.height=5}

print(paste("For creating the QC PCA plots",  scale_used, "scaling was used"))

print(paste("PCA plot using the",  length(which(ltr_rsd$RSD < 30)), "lipids in the final dataset" ))
pca_p[[1]][[1]]


print(paste("PCA plot using lipid class created by summing individual lipids"))
pca_p[[2]][[1]]

```

