---
title: "ANPC SkylineR and Lipid_exploreR notebook"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
This notebook is designed for use with the ANPC targeted lipid method. 
Section 1: SkylineR is designed to optimise lipidomics data processing in combination with skyline.
Section2: Lipid_exploreR is designed to explore, visualise and QC check the data.

The sections should be run in sequence. However should section 1 already be completed, section 2 can be run independently at a later date.

------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
Section 1 - SkylineR

This notebook is designed to optimise lipidomics data processing in combination with skyline.

It will perform:
- retention time optimisation using LTR QC
- peak boundary fitting to all samples

REQUIREMENTS:
- A subfolder containig mzML files. Proteowizard should be used to convert Sciex targeted lipidomics data using proteowizard default settings
- Filename should match the LIMS where possible
- mzML files from LTR samples must have "LTR" in their filename
- A csv template containing the target transition details. ONLY the following column headers should be present: 
      - "Molecule List" (lipid family (e.g. CE))
      - "Precursor Name" (lipid name (e.g. CE(14:0)))
      - "Precursor Mz" (e.g. 614.6)
      - "Precursor Charge" (set as 1)
      - "Product Mz" (e.g. 369.4)
      - "Product Charge" (set as 1)
      - "Explicit Retention Time" (e.g. 11.66)
      - "Explicit Retention Time Window" (leave at 0.5)
      - "Note" in the column "Note" insert the SIL IS to be used for the target lipid. For the rows containing SIL IS themselves leave the note column blank.



```{r, skylineR,  eval = FALSE, echo = FALSE, results = FALSE, warning=FALSE, message=FALSE}
package_list <- c("plyr", "tidyverse", "janitor", "httr")
loaded_packages <- lapply(package_list, require, character.only = TRUE)
rm(loaded_packages)

# Source R script from Github
skylineR_script <- GET(url = "https://raw.githubusercontent.com/lukewhiley/metabolomics_code/main/ANPC_lipidomics_tools/scripts/2021-LGW-lipidomics-skylineR_script.r") %>% content(as = "text")

eval(parse(text = skylineR_script), envir = .GlobalEnv)


```



-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

Section 2 - lipid_exploreR

This notebook is designed to explore the dataset and QC check it. The script generates a report and a final dataset that can be used for data modeling.

```{r, lipid_exploreR,  eval = FALSE, echo = FALSE, results = FALSE, warning=FALSE, message=FALSE}
package_list <- c("plyr", "tidyverse", "janitor", "gridExtra", "ggpubr", "readxl", "cowplot", "scales", "stats", "devtools", "metabom8", "shiny", "plotly", "svDialogs", "DataEditR", "htmlwidgets", "httr", "htmlTable", 'statTarget')
loaded_packages <- lapply(package_list, require, character.only = TRUE)
rm(loaded_packages, package_list)

#lipidomics_functions_script <- GET(url = "https://raw.githubusercontent.com/lukewhiley/metabolomics_code/main/ANPC_lipidomics_tools/functions/LGW_function_lipidomics_tools.r") %>% content(as = "text")
#eval(parse(text = lipidomics_functions_script), envir = .GlobalEnv)
#rm(lipidomics_functions_script)

# Source R script from Github and prepare data - imports data from section 1. Lipidomics peak picking from skyline
lipid_exploreR_script <- GET(url = "https://raw.githubusercontent.com/lukewhiley/metabolomics_code/main/ANPC_lipidomics_tools/scripts/2021-LGW-lipidomics-lipid_exploreR_script.r") %>% content(as = "text")
eval(parse(text = lipid_exploreR_script), envir = .GlobalEnv)
rm(lipid_exploreR_script)

# produce a final report
fileConn<-file(paste(project_dir_html, "/ANPC_lipidomics-report-generatoR.r", sep=""))
writeLines(GET("https://raw.githubusercontent.com/lukewhiley/metabolomics_code/main/ANPC_lipidomics_tools/scripts/2021-LGW-lipidomics-report_generatoR_script.R") %>% 
  content(as = "text"), fileConn)
close(fileConn)

rmarkdown::render(input = paste(project_dir_html, "/ANPC_lipidomics-report-generatoR.r", sep=""),
                  output_format = "html_document",
                  output_dir = paste(project_dir_html),
                  output_file = paste(project_name, "_", user_name, "_lipid_exploreR_QC_report.html", sep="")
                  )

file.remove(paste(paste(project_dir_html, "/ANPC_lipidomics-report-generatoR.r", sep="")))

browseURL(paste(project_dir_html, "/",project_name, "_", user_name, "_lipid_exploreR_QC_report.html", sep="")) #open plotly widget in internet browser

write_csv(final_corrected_data, 
          paste(project_dir, "/", project_name, "_", user_name, "_processed_lipid_data.csv", sep="")
          )

write_csv(final_corrected_class_lipid_data, 
          paste(project_dir, "/", project_name, "_", user_name, "_processed_lipid_class_data.csv", sep="")
          )

dlg_message("Lipidomics pre-processing and QC complete. Thanks for using skylineR and lipid exploreR. TTFN.", type = 'ok')
dlg_message("PS. an official html report can be found in the project folder", type = 'ok')


```










